---
title: "goalie_saves"
author: "sam toole"
date: "2024-08-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#install.packages(dplyr)
library(dplyr)
```


This chunk reads a table from FBref with stats on goalies from the 'Big 5' Leagues
```{r}
GK=read_html('https://fbref.com/en/comps/Big5/keepers/players/Big-5-European-Leagues-Stats',as.data.frame=TRUE,stringAsFactors=TRUE) %>% html_nodes("table") %>% .[[1]] %>% html_table(fill=TRUE)
```

This cleans the data, converts the column types to numeric, and adds a few additional stats 
```{r}

names(GK) <- GK[1,]
GK <- GK[-1,] 
GK <- GK[,-1] 
GK <- GK[, -((ncol(GK)-1):ncol(GK))]

GK <- GK %>%
  mutate(
    MP = as.numeric(MP),
    Starts = as.numeric(Starts),
    Min = as.numeric(Min),
    `90s` = as.numeric(`90s`),
    GA = as.numeric(GA),
    GA90 = as.numeric(GA90),
    SoTA = as.numeric(SoTA),
    Saves = as.numeric(Saves),
    `Save%` = as.numeric(`Save%`),
    W = as.numeric(W),
    D = as.numeric(D),
    L = as.numeric(L),
    CS = as.numeric(CS),
    `CS%` = as.numeric(`CS%`),
    PKatt = as.numeric(PKatt),
    PKA = as.numeric(PKA),
    PKsv = as.numeric(PKsv),
    PKm = as.numeric(PKm),
  )
  
GK$SoTA90 <- GK$SoTA / GK$`90s`
GK$exSaves <- GK$SoTA90 * (GK$`Save%`/100)

```

Specifing the model 
```{r, ECHO = FALSE, include=FALSE}

model_saving <- lm(exSaves ~ MP + Starts + Min + `90s` + GA + GA90 + SoTA + 
                     `Save%` + W + D + L + CS + `CS%` + PKatt + PKA + PKsv + 
                     PKm + `Save%` + SoTA90, data = GK)
summary(model_saving)
```


Insert keeper name and then run chunk to estimate saves using the model
```{r}
goalie <- GK %>% filter(Player == "Guglielmo Vicario")

predicted_saves_goalkeeper1 <- predict(model_saving, newdata = goalie)
print(predicted_saves_goalkeeper1)
```


And here is a similar model to predict how many saved shots a team may get 
```{r}
tmshots=read_html('https://fbref.com/en/comps/Big5/shooting/squads/Big-5-European-Leagues-Stats',as.data.frame=TRUE,stringAsFactors=TRUE) %>% html_nodes("table") %>% .[[1]] %>% html_table(fill=TRUE)
names(tmshots) <- tmshots[1,]
tmshots <- tmshots[-1,] 

numeric_columns <- tmshots[sapply(tmshots, is.numeric)]
numeric_columns <- na.omit(numeric_columns)

tmshots$glsprgm <- as.numeric(tmshots$Gls) / as.numeric(tmshots$`90s`)
tmshots$exsvs <-  as.numeric(tmshots$`SoT/90`) - as.numeric(tmshots$glsprgm)

model_shooting <- lm(exsvs ~ `Sh/90` + `SoT/90` + 
                       `G/Sh` + `G/SoT` + xG + `npxG/Sh`, data = tmshots)

```

Input the team name here to get an estimation of how many saves a keeper will get playing against them. 
```{r}
opshots <- tmshots %>% filter(Squad == "Nott'ham Forest")
predicted_shots_saved <- predict(model_shooting, newdata = opshots)

unname(predicted_shots_saved)
```